Ownership types in theory and practice (in Rust)

### 2023-01-22 paper/paper.tex
Some progress on slides

There are a variety of potential occurrences of memory management issues.
Examples include \emph{buffer overflows}, where data is accessed outside the bounds of an array, \emph{memory leaks}, where the programmer forgets to free memory that has been allocated, causing the program to, eventually, run out of memory, \emph{Use-After-Free}, where memory is accessed after it has been freed,\emph{Double-Free}, where memory is freed again after it had already been freed before, and the use of uninitialized memory.
These issues can not only enable potential malicious exploits, they can also result in incorrect program results, the decrease of a program's performance or seemingly random program crashes.

@@

\begin{lstlisting}[language=Rust]
let mut v = vec![10, 11];
let vptr = &mut v[1]; // points *into* v
v.push(12);
println!("{}", *vptr); // compiler error
\end{lstlisting}

Syntactically, the C++ and Rust versions are very similar.
Notably, the \emph{push} method in line 3 has the following signature: \verb|pub fn push(&mut self, value: T)| \cite{rust-vector-documentation}.
While the C++ version will compile successfully, causing a run-time security vulnerability, the Rust compiler shows an error: "Cannot borrow \emph{v} as mutable more than once at a time."
For now, it suffices to say that the Rust compiler disallows both lines 2 and 3 to create mutable references to \emph{v} if one of the mutable references is used later (like \emph{vptr} is in line 4).
In \ref{sec:rust-references}, we will explain in more detail how the Rust borrow-checker is able to detect this problem using Rust's ownership rules.

@@

\subsection{Lifetimes}

@@

\subsection{Other ownership implementations}

>>>>

There are a variety of potential occurrences of memory management issues.
Examples include \emph{buffer overflows}, where data is accessed outside the bounds of an array, \emph{memory leaks}, where the programmer forgets to free memory that has been allocated, causing the program to, eventually, run out of memory, \emph{Use-After-Free}, where memory is accessed after it has been freed, \emph{Double-Free}, where memory is freed again after it had already been freed before, and the use of uninitialized memory.
These issues can not only enable potential malicious exploits, they can also result in incorrect program results, the decrease of a program's performance or seemingly random program crashes.

@@

\begin{lstlisting}[language=Rust]
let mut v = vec![10, 11];
let vptr = &mut v[1]; // points into v
v.push(12);
println!("{}", *vptr); // compiler error
\end{lstlisting}

Syntactically, the C++ and Rust versions are very similar.
Notably, the \emph{push} method in line 3 has the following signature: \verb|pub fn push(&mut self, value: T)| \cite{rust-vector-documentation}.
While the C++ version will compile successfully, causing a run-time security vulnerability, the Rust compiler shows an error: "Cannot borrow \emph{v} as mutable more than once at a time."
For now, it suffices to say that the Rust compiler disallows both lines 2 and 3 to create mutable references to \emph{v} if one of the mutable references is used later (like \emph{vptr} is in line 4).
In \ref{sec:rust-references} and \ref{sec:rust-lifetimes}, we will explain in more detail how the Rust borrow-checker is able to detect this problem using Rust's ownership rules.

@@

\subsection{Lifetimes}
\label{sec:rust-lifetimes}

@@

\subsection{Other ownership implementations}
\label{sec:other-ownership-implementations}

### 2023-01-22 slides/slides.tex
Some progress on slides

\begin{frame}[fragile]{Memory safety}

C++
\begin{lstlisting}[language=C++]
std::vector<int> v { 10, 11 };
int *vptr = &v[1];  // points into v
v.push_back(12);
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

Rust
\begin{lstlisting}[language=Rust]
let mut v = vec![10, 11];
let vptr = &mut v[1];  // points *into* v
v.push(12);
println!("{}", *vptr); // compiler error
\end{lstlisting}

\end{frame}

\section{Background}

\subsection{Ownership types in theory}

\begin{frame}{Domain specific terms}
\end{frame}

\begin{frame}{Owners-as-dominators}

@@

  \caption{Owners-as-dominators: Solid lines indicate "owns" and dotted lines indicate "references".}
\end{figure}

\end{frame}

\begin{frame}{Owners-as-modifiers}

@@

    \draw[->,thick,dashed,cgreen] (Car) edge[bend left] node[right] {owner} (Engine);
    \draw[->,thick,dashed,cgreen] (Engine) edge[bend left] node[left] {inside} (Car);
    \draw[->,thick,dashed,cgreen] (Car) edge node[above] {sibling} (Driver);
    \draw[->,thick,dashed,orange] (World) edge[bend right=90] node[left] {read only} (Engine);
    \draw[->,thick,dashed,orange] (Driver) edge[bend left] node[right] {read only} (Engine);
  \end{tikzpicture}

  \caption{Owners-as-modifiers: Solid lines indicate "owns" and dotted lines indicate "references".}
\end{figure}

\end{frame}

\section{Concept}

\subsection{Ownership in Rust}

\begin{frame}{Memory allocation system}

\end{frame}

@@

\section{Evaluation}

\subsection{Another ownership implementation}

>>>>

\begin{frame}{Aliasing}
\begin{itemize}
  \item object aliasing = accessing the same memory through different symbolic names
  % TODO: example
  \item unintentional aliases $\rightarrow$ bugs, hard to detect
  \item ownership types goal: restrict object aliasing (through rules) $\rightarrow$ improve memory safety
\end{itemize}
\end{frame}

\begin{frame}{Memory safety}
\begin{itemize}
  \item ownership types goal: improve memory safety
  \item Google and Microsoft: 70 percent of vulnerabilities due to memory safety issues
  \item National Security Agency (USA) recommends memory-safe languages (instead of C/C++)
  \item examples: buffer overflows, memory leaks, double-free, use-after-free
  \item result: malicious exploits, incorrect program results, "random" crashes
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Memory safety: example}
C++
\begin{lstlisting}[language=C++]
std::vector<int> v { 10, 11 };
int *vptr = &v[1];  // points into v
v.push_back(12);    // v buffer reallocated => vptr dangling
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

Rust
\begin{lstlisting}[language=Rust]
let mut v = vec![10, 11];
let vptr = &mut v[1];  // 1. mutable reference
Vec::push(&mut v, 12); // 2. mutable reference
println!("{}", *vptr); // compiler error
\end{lstlisting}
\end{frame}

\section{Ownership types in theory}

\begin{frame}{Ownership types}
\begin{itemize}
  \item object-oriented programs:
  \begin{itemize}
    \item objects can reference any other object
    \item objects can read and modify other objects' fields
    \item makes it difficult to understand, to maintain, and to reason about
  \end{itemize}
  \item using ownership types:
  \begin{itemize}
    \item limit which objects can be referenced
    \item can specify whether the referenced objects may be mutated or just read from
  \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Ownership types: core concepts}
\begin{itemize}
  % TODO: different highlighting than italics
  \item objects can be \emph{owners} of other objects
  \item \emph{inside} relation: \emph{a} is \emph{inside} \emph{b} if they are the same object or \emph{a} is transitively owned by \emph{b}
  \item \emph{outside} relation: converse of \emph{inside} relation
  \item \emph{ownership context}: set of all objects that an object owns
  \item \emph{siblings}: objects in the same ownership context
  \item special \emph{world} object is root of the ownership hierarchy tree
\end{itemize}

% TODO: make spacing between vertical objects smaller
\begin{figure}[fragile]
  \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
    \node[object] (World) {world};
    \node[object] (Car) [below=of World] {car};
    \node[object] (Driver) [right=of Car] {driver};
    \node[object] (Engine) [below=of Car] {engine};

    \draw[->,thick] (World) to (Car);
    \draw[->,thick] (World) to (Driver);
    \draw[->,thick] (Car) to (Engine);
  \end{tikzpicture}

  \caption{solid lines indicate "owns"}
\end{figure}
\end{frame}

\begin{frame}{Owners-as-dominators}
For \emph{a} to validly reference \emph{b}:
\begin{enumerate}
  \item \emph{a} is the owner of \emph{b},
  \item \emph{a} and \emph{b} are siblings, or
  \item \emph{b} is outside of \emph{a}.
\end{enumerate}

@@

  \caption{solid lines indicate "owns" and dotted lines indicate "references"}
\end{figure}
\end{frame}

\begin{frame}{Owners-as-dominators: implications}
\begin{itemize}
  \item any external reference to an object must go through its owner
  \item objects are only protected from external access, not internal access
  \item owners-as-dominators: owner is dominator for all of its owned objects
  \item doesn't allow common idioms that involve aliasing
  \item model restricts \emph{where} references can point, not \emph{how} references are used
\end{itemize}

    \draw[->,thick,dashed,cgreen] (Car) edge[bend left] node[right] {owner} (Engine);
    \draw[->,thick,dashed,cgreen] (Engine) edge[bend left] node[left] {inside} (Car);
    \draw[->,thick,dashed,cgreen] (Car) edge node[above] {sibling} (Driver);
    \draw[->,thick,dashed,red] (World) edge[bend right=90] node[left] {invalid} (Engine);
    \draw[->,thick,dashed,red] (Driver) edge[bend left] node[right] {invalid} (Engine);
  \end{tikzpicture}
\end{figure}
\end{frame}

\begin{frame}{Owners-as-modifiers}
Same rules as owners-as-dominators, but \emph{a} can also reference \emph{b} through \emph{r}, if:
\begin{enumerate}
  \item \emph{r} is a read-only reference and only pure methods (that don't modify existing objects) can be called on it.
\end{enumerate}

@@

    \draw[->,thick,dashed,cgreen] (Car) edge[bend left] node[right] {owner} (Engine);
    \draw[->,thick,dashed,cgreen] (Engine) edge[bend left] node[left] {inside} (Car);
    \draw[->,thick,dashed,cgreen] (Car) edge node[above] {sibling} (Driver);
    \draw[->,thick,dashed,orange] (World) edge[bend right=90] node[left] {read only} (Engine);
    \draw[->,thick,dashed,orange] (Driver) edge[bend left] node[right] {read only} (Engine);
  \end{tikzpicture}
\end{figure}
\end{frame}

\begin{frame}{Owners-as-modifiers: implications}
\begin{itemize}
  % \item any external reference to an object must go through its owner
  % \item objects are only protected from external access, not internal access
  \item owners-as-modifiers: only the owners can modify objects
  % \item doesn't allow common idioms that involve aliasing
  % \item model restricts \emph{where} references can point, not \emph{how} references are used
\end{itemize}

    \draw[->,thick,dashed,cgreen] (Car) edge[bend left] node[right] {owner} (Engine);
    \draw[->,thick,dashed,cgreen] (Engine) edge[bend left] node[left] {inside} (Car);
    \draw[->,thick,dashed,cgreen] (Car) edge node[above] {sibling} (Driver);
    \draw[->,thick,dashed,orange] (World) edge[bend right=90] node[left] {read only} (Engine);
    \draw[->,thick,dashed,orange] (Driver) edge[bend left] node[right] {read only} (Engine);
  \end{tikzpicture}
\end{figure}
\end{frame}

\begin{frame}{From theory to Rust}
\begin{itemize}
\end{itemize}
\end{frame}

\section{Ownership in Rust}

\begin{frame}{Memory allocation system}
\end{frame}

@@

\section{Evaluation: another implementation}

% \subsection{Another ownership implementation}
